---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---
```{r echo=FALSE, results="hide"}
library(splines)
library(modelr)
```

```{r}
data <- read_csv("SAT_School_Participation_and_Performance__2012-2013.csv")
```

```{r}
#Alex's contribution: Tidying up the data
df <- data %>% select(-1, -6, -9, -12) %>% rename(district = "District", school = "School", t_takes2012 = "Test-takers: 2012", t_takes2013 = "Test-takers: 2013", part_rate2012 = "Participation Rate (estimate): 2012", part_rate2013 = "Participation Rate (estimate): 2013", perc_mb2012 = "Percent Meeting Benchmark: 2012", perc_mb2013 = "Percent Meeting Benchmark: 2013")
df <- df %>% dplyr::filter(!(is.na(t_takes2012) | is.na(t_takes2013) | is.na(part_rate2012) | is.na(part_rate2013) | is.na(perc_mb2012) | is.na(perc_mb2013)))
df
```


## Introduction
In this project, the dataset provides Benchmark Meeting and participation rate, but it did not show what exactly how many people reach the Benchmark in that year, and the Percent Meeting Benchmark of the total number of test takers. Therefore, we created a  new index call BMR(Benchmark Meeting Rate),we find all the seniors in a school fist, then calculate BMR through Benchmark-meeting seniors divided by the number of total seniors. We use BMR to evaluate the quality of education in the districts. This report we will analyze the distribution of Benchmark Performance in 2012 and 2013.

bmr = number of meeting Benchmark / number of total seniors = (t_takes*perc_mb) / (t_takes/part_rate) = pec_mb*part_rate
bmr = perc_mb*part_rate*0.0001


We use bmr because it's a better measurement for comparing how well schools do. If 2 schools have the same percentage meeting benchmark, but one of them has a higher participation rate then the one with the higher participation rate is the better school.

## Questions and Findings

```{r}
#Alex's contribution: creating BMR formula
#df1 is for testtakers for each school+year
df1 <- df %>% 
  select(1:4) %>% 
  rename(`2012` = t_takes2012, `2013` = t_takes2013) %>% 
  gather(3,4,key = "year", value = "t_takes") %>% 
  arrange(school)

#df2 is participation rate for each school+year
df2 <- df %>% select(1,2,5,6) %>% 
  rename(`2012` = part_rate2012, `2013` = part_rate2013) %>% 
  gather(3,4,key = "year", value = "part_rate")

#df3 is percentage meeting benchmark for each school+year
df3 <- df %>% 
  select(1,2,7,8) %>% 
  rename(`2012` = perc_mb2012, `2013` = perc_mb2013) %>% 
  gather(3,4,key = "year", value = "perc_mb")

#df4 combines them all
df4 <- df1 %>% 
  full_join(df2,by = c("district","school","year")) %>% 
  full_join(df3,by = c("district","school","year"))
df4 <- df4 %>% 
  mutate(bmr = perc_mb*part_rate*1e-4)
```



First we'll get the senior population for each school (denoted as pop)
```{r}
data <- df4 %>% mutate(pop = floor(1e2*t_takes / part_rate))
data
```

Now lets plot it
```{r}
ggplot(data) + 
  geom_point(aes(pop,bmr),alpha=2/5,size=3.5) + 
  facet_wrap(~year) + 
  theme_bw() + 
  labs(title="Senior Population vs Benchmark Meeting Rate",caption="This shows the population vs bmr for each year.")
```


The data is relatively scattered, but we can see a weak positive linear trend.

Let's use mean-square residuals
```{r}
#mean-square residuals
measure_distance <- function(mod,data){
  diff <- data$bmr - (mod[1] + data$pop*mod[2])
  sqrt(mean(diff^2))
}

best <- optim(c(0, 0), measure_distance, data = data)

ggplot(data, aes(pop, bmr)) + 
  geom_point(size = 2, colour = "grey30") + 
  geom_abline(color="blue",intercept = best$par[1], slope = best$par[2]) + 
  theme_bw()
```


```{r}
mod1 <- lm(bmr ~ ns(pop, 1), data = data)
mod2 <- lm(bmr ~ ns(pop, 2), data = data)
mod3 <- lm(bmr ~ log(pop, base = exp(1)), data = data)
mod4 <- lm(bmr ~ I(pop^2), data = data)

data %>% 
  gather_predictions(mod1, mod2, mod3, mod4) %>%
  ggplot(aes(pop, bmr)) + 
  geom_point(alpha=2/5) +
  geom_line(aes(pop,pred), colour = "red") +
  facet_grid(year~ model) + 
  theme_bw()
```

Let's check the residuals for any patterns
```{r}
data %>%
  gather_residuals(mod1,mod2,mod3,mod4) %>%
  ggplot(aes(resid)) + 
  geom_freqpoly(binwidth = 0.05) + 
  geom_vline(xintercept = 0, colour = "Green", size=0.5) + 
  facet_grid(year ~ model)
```

Looks approximately normal for all.

```{r}
data %>%
  gather_residuals(mod1,mod2,mod3,mod4) %>%
  ggplot(aes(pop, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 2) + 
  geom_point(alpha=2/5) + 
  facet_grid(year ~ model)
```

Coefficient of Determination (r^2)

```{r}
summary(mod1)$r.squared
summary(mod2)$r.squared
summary(mod3)$r.squared
summary(mod4)$r.squared
```

These coefficients SUCK

We find the highest participation rate of 2012 and 2013.
```{r}
pr2012_highest <- data %>%
    dplyr::filter(part_rate ==100, year == 2012) %>%
    select(district) 
ggplot(pr2012_highest) +
  geom_bar(aes(district))+
coord_flip()
```

```{r}

pr2013_highest <- data %>%
    dplyr::filter(part_rate == 100, year == 2013) %>%
    select(district) 

ggplot(pr2013_highest) +
  geom_bar(aes(district))+
coord_flip()
```

## Conclusion
According to the graphs, we removed all the bull values in oth 2012 and 2013from data of SAT School Participation and Performance informations in Connecticut state， there are around 170 schools remaining. From the graph of compare to population and BMR, we find that students number in most school is less than 300, and most of them lower than 0.5 BMR. Population of 2013 increase a little bit than 2012, but it did not show big change overall. There are 3 schools from three districts have the highest BMR in both years, which is "Darien High School", "New canaan High School" and "Wilton High School". All three school have student scale around 300. We concluded that in the state of Connecticut, student scale around 300 can make best Benchmark Performance.

## Contributions
Alex - Created the formula for BMR, came up with the ideas on what to explain from our model.
Michael - Michael created the models for the graphs of senior population vs bmr, the residual plots,
Hongyang - Wrote the Introduction and Conlcusion, added the graphs for the highest participation rates.

