---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
---
title: "R Notebook"
output: html_notebook
---

```{r echo=FALSE, results="hide"}
library(splines)
library(modelr)
library(modelr)
library(tidyverse)
```


 What was the trend of world happiness scores from 2015 to 2019?
I combined the all data of for 5 years, and added Region for the last three years.
```{r}
aaa <- read_csv("D:\\R\\2015.csv")
Country_Region <- aaa %>%
  select(Country, Region)

da15 <- read_csv("D:\\R\\2015.csv")  
rank2015 <- da15 %>%
 select(Country,Region,`Happiness Rank`,`Happiness Score`) %>%
 rename(Rank=`Happiness Rank`) %>%
 rename(Score=`Happiness Score`) %>%
 mutate(year="2015")

da16 <- read_csv("D:\\R\\2016.csv")
rank2016 <- da16 %>%
 select(Country,Region,`Happiness Rank`,`Happiness Score`) %>%
 rename(Rank=`Happiness Rank`) %>%
 rename(Score=`Happiness Score`) %>%
 mutate(year="2016")

da17 <- read_csv("D:\\R\\2017.csv")
rank2017 <- da17 %>%
 left_join(Country_Region,by = "Country")%>%
 select(Country,Region,Happiness.Rank,Happiness.Score) %>%
 rename(Rank=Happiness.Rank) %>%
 rename(Score=Happiness.Score) %>%
 mutate(year="2017")

da18 <- read_csv("D:\\R\\2018.csv")
rank2018 <- da18 %>%
 rename(Country=`Country or region`) %>%
 rename(Rank=`Overall rank`) %>%
left_join(Country_Region,by = "Country")%>%
 select(Country,Region,Rank, Score) %>%
  mutate(year="2018")

da19 <- read_csv("D:\\R\\2019.csv")
rank2019 <- da19 %>%
 rename(Country=`Country or region`) %>%
 rename(Rank=`Overall rank`) %>%
left_join(Country_Region,by = "Country")%>%
 select(Country,Region,Rank, Score) %>%
 mutate(year="2019")

total<- tibble(
  Country=c(rank2015$Country ,rank2016$Country ,rank2017$Country  ,rank2018$Country ,rank2019$Country),
  Region=c(rank2015$Region ,rank2016$Region ,rank2017$Region  ,rank2018$Region ,rank2019$Region),
  Rank=c(rank2015$Rank ,rank2016$Rank ,rank2017$Rank ,rank2018$Rank ,rank2019$Rank),
  Score=c(rank2015$Score ,rank2016$Score ,rank2017$Score ,rank2018$Score ,rank2019$Score),
  year=c(rank2015$year ,rank2016$year ,rank2017$year ,rank2018$year ,rank2019$year)
  ) %>%
arrange(Country,Rank)

totl <- na.omit(total)  
```

The geom_line graph of happiness score.
```{r}
totl %>% 
  ggplot(aes(year, Score, group = Country)) +
    geom_line(alpha = 1/3)
```

Get the data of predictions and residuals.
```{r}
by_country <- totl%>% 
  group_by(Country) %>% 
  nest()

country_model <- function(df) {
  lm(Score ~ year, data = totl)
}

by_country <- by_country %>% 
  mutate(model = map(by_country$data, country_model))


by_country <- by_country %>% 
  mutate(
    pred = map2(data, model, add_predictions)
  )
preds <- unnest(by_country, pred)

by_country <- by_country %>% 
  mutate(
    resids = map2(data, model, add_residuals)
  )
resids <- unnest(by_country, resids)

resids
```


The boxplot graph of happiness score
```{r}
ggplot(resids, aes(year, Score)) +
 geom_boxplot()
```

The graph of residuals.
```{r}
ggplot(resids, aes(year, resid)) +
 geom_boxplot()
```
The graph of predictions.
```{r}
ggplot(preds, aes(year, pred)) +
 geom_boxplot()
```

Then I separate the countries into different Regions to check each Continent's happiness score Continent.
```{r}
ggplot(resids, aes(year, Score)) +
 geom_boxplot() +
 facet_wrap(~ Region, nrow = 2)

```

The geom_line of predictions 
```{r}
preds %>% 
  ggplot(aes(year, pred)) +
    geom_line(aes(group = Country), alpha = 1/3) + 
    geom_smooth(se = FALSE)
```

